Osa

### 2021-09-08 article/article.Rmd
polished intro for now

---
title: "Osa"
output: 
  html_notebook: 
    toc: yes
    code_folding: hide
    theme: spacelab
    fig_caption: yes
---

@@

1.  importance of reforestation

@@

@@

To realize this potential, it is fundamental that we better understand how general ecological processes manifest under widespread conservation contexts.

@@

Most tropical forests are secondary (FAO 2020), meaning that forest regeneration is widespread.

@@

Still, forests provide key functions like wood, carbon storage, and biodiversity maintenance.

Therefore, better understanding forest regeneration is an integral part of achieving more sustainable development.

Some general patterns of forest regeneration have emerged, but this also includes the lack of pattern in some cases.

@@

@@

@@

1.  shade x distance + timber sp

1.  biomass x distance + timber sp

    -   can break down biomass by component

@@

@@

-   Neighborhood influence metric (Lorimer 1983, applied in Looney et al 2016)
-   Data and analytical code available on GitHub.../DRYAD

>>>>

---
title: "Osa"
output: 
  html_notebook: 
    toc: yes
    code_folding: hide
    theme: spacelab
    fig_caption: yes
bibliography: references.bib
---

@@

1.  importance of reforestation/regeneration/regrowth

@@

    2.  How does the forest regenerate by distance to primary forest?

        1.  how much [effort] do we need to re-seed? / how much can we leave to nature?

Global forest and landscape regeneration is a key conservation and climate change mitigation strategy, adopted internationally (\@UN14 Paris, Bonn Challenge).

-   Tropical forests store most terrestrial biomass @pan2013 and biodiversity (CITE, but are losing functions, especially in the neotropics @hubau2020, as most tropical forests become secondary (\@FAO20.

Secondary forests can regenerate relatively quickly compared to old-growth forests in some ways, but in other ways remain distinct.

-   Neotropical secondary forests accumulate biomass \>10x faster than old-growth forests @poorter2016, and offset \~10% of carbon emissions from Amazonian deforestation @smith2020 .
-   They also recover taxonomic richness quite quickly, averaging 80% of old-growth forest levels in 20y @rozendaal2019 .
-   However, secondary neotropical forests diverge widely in taxonomic composition @norden2015 .

Divergent communities among secondary forests may warrant more localized conservation efforts, studying relative effects of generalized community processes.

Half of all forests lie half of a kilometer from an ecotone @haddad2015, suggesting that edge effects from adjacent habitats are widespread aspects of secondary forest regeneration.

-   Tropical forests lose 22% carbon along edges \>100m, in part due to shorter trees and less leaf mass @ordway2020 .

Here we report the edge effects of an abandoned wet neotropical timber plantation.

-   We hypothesized that remnant timber trees would mediate forest regrowth by maintaining uniform shade that would suppress shade-intolerant / pioneer trees / taxa and allow for more diversity.

@@

@@

@@

@@

Edge effects are known for degrading primary forest, but they can also be used to assist reforestation.

Traditional edge effects act by mortality of large trees that makes room for pioneer species to enter, but existing interior pioneer species also have dispersal radii that overlap with the adjacent exterior, potentially facilitating regeneration.

Encroachment vs. facilitation relies on a balance between mortality (per area) and dispersal + establishment + stem growth.

So pioneer species can be helpful for reforestation though they are typically seen as bad for old forest conservation.

@@

-   how can we expect it to regrow?
-   how far/much [effort] do we need to re-seed?

@@

1.  fig1

    1.  shade x distance @ordway2020 + timber sp?

    2.  height @ordway2020

    3.  wood density @poorter2019

    4.  total biomass x dist @ordway2020 + timber sp?

@@

1.  fig2

    1.  richness

@@

-   Neighborhood influence metric (Lorimer 1983, applied in Looney et al 2016)

-   Data and analytical code available on GitHub.../DRYAD
